M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2013-02-25 notes/thesis-latex-template/thesis.tex
Update thesis.

When $\beta$ becomes large, the salt concentration $C$ near the front of the particle 
(at $\theta \approx 0$) approaches zero, while creating a wake behind the particle 
(at $\theta \approx \pi$).
This results in high gradients in $\log C$ near the front of the particle (Figures \ref{fig:LargeBeta_C1}, \ref{fig:LargeBeta_C2} and
\ref{fig:LargeBeta_C3}) and, therefore, a strong electric field $\bE$ forms near the particle surface (Figures \ref{fig:LargeBeta_Phi1}, \ref{fig:LargeBeta_Phi2} and \ref{fig:LargeBeta_Phi3}) due to effective boundary conditions in (\ref{eq:ephor_bnd}). 
The streamline pattern does not change drastically for different electric field magnitudes. 
The resulting flow pattern is similar to the Stokes creeping flow, as shown in Figure
\ref{fig:LargeBeta_Psi}.

The numerical results suggest that a boundary 
layer starts to form around the front of the particle for strong
electric fields (see Figure \ref{fig:LargeBeta_BoundaryLayer}). 
This causes numerical problems to the current solver -- yielding convergence
problems for large $\beta$. The convergence problems start earlier
for smaller grid size, and for higher $R_{max}$ values, probably
due to insufficient resolution at the $\brhat$ direction near
the particle surface.

\begin{figure}
    \begin{center}
    \includegraphics[width=.45\textwidth]{figs/LargeBetaV_[128_256_512]_Rmax=10.eps}
    \includegraphics[width=.45\textwidth]{figs/LargeBetaV_[512x512]_Rmax=10_30_100.eps}
        \caption[Large $\beta$ results for $\cU$]
{Large $\beta$ results for $\cU$ ($\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$). 
The left figure shows the numerical results of $\cU(\beta)$ for $R_{max} = 10$ and the following grid sizes: $128 \times 128$, $256 \times 256$, $512 \times 512$. 
The right figure shows the numerical results of $\cU(\beta)$ for
$512 \times 512$ grid for $R_{max} = 100, 30, 10$. For small values of $\beta$, the linear
regime of $\cU(\beta)$ is dominant. However, for larger values of $\beta$, the numerical solution
deviates significantly from the linear regime. For $\beta > 6$, the numerical solver convergence
is slowing and for $\beta \sim 7$, the solver fails to converge. As the grid resolution near
the particle surface becomes finer (when $N_r$ increases or $R_{max}$ decreases), the solver
convergence is better.}
	    \label{fig:LargeBetaV_grids}	    
    \end{center}
\end{figure}
\begin{figure}
    \begin{center}
    \includegraphics[width=.5\textwidth]{figs/LargeBetaRatio.eps}
        \caption[Rate of convergence estimation for $\cU$ at large $\beta$]
        {Rate of convergence estimation for $\cU$ at large $\beta$  
        ($\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$). 
        The numerical results of $\cU(\beta)$, for the grid sizes of 
        $128 \times 128$, $256 \times 256$, $512 \times 512$. 
        The rate of convergence $r(\beta)$ is estimated by 
        $r(\beta) \triangleq \left(\cU^{4h}(\beta) - \cU^{2h}(\beta)\right)/\left(\cU^{2h}(\beta) - \cU^{h}(\beta)\right)$.
        For small $\beta$, $r(\beta) \rightarrow 4$, showing 
        quadratic convergence for the linear regime. 
        For moderate $\beta$, the convergence rate slows down 
        and deteriorates for large $\beta$ values.
        }
	    \label{fig:LargeBetaRatio}
    \end{center}
\end{figure}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=1\textwidth]{figs/BoundaryLayerWidth.eps}
        \caption[Electric field boundary layer]
        {Electric field boundary layer formation: grid size $512 \times 512$, 
        $R_{max} = 30$, $\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$. The left
        graph shows normalized electric field radial component $E_r$ near the particle front
        ($\theta \approx 0$). The numerical results for $e(\rho) = E_r(r=1+\rho, \theta=0)$
        are normalized by $\hat{e}(\rho) = \left(e(\rho) - \beta\right)/\left(e(0) - \beta\right)$,
        where $e(\rho = \infty) = \beta$.
        The normalized electric field satisfies $\hat{e}(\rho) \in [0, 1]$ and is shown as function
        of $\rho = r - 1$ at the left graph.
        In order to estimate the width of the boundary layer, for each $\beta$, we compute the value 
        of $\bar\rho(\beta) = \min_\rho \{ \rho : \hat{e}(\rho) < 0.5\}$. Each such $\bar\rho(\beta)$
        is plotted at the left graph as a solid dot on the corresponding line.
        In addition, $\bar\rho(\beta)$ is plotted as a function of $\beta$ using logarithmic 
        $y$ axis at the right graph to show that $\bar\rho(\beta) \propto e^{-A\beta}$.
        Thus, a boundary layer is indeed forming near the front of the particle.
        }
	    \label{fig:LargeBeta_BoundaryLayer}	    
    \end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\section{Discussion} \label{sec:discussion}
A numerical framework for solver construction for a
nonlinear electrokinetic transport system is described.
The framework is applied to two electrokinetic problems: the migration of ion-exchangers
and surface-charged inert particle electrophoresis.
The numerical results exhibit a good correspondence to the asymptotic linear model as
shown above.
Moreover, two interesting phenomena (cubic regime and downstream vortex) were predicted
using the numerical results, and were verified using asymptotic expansion of the
nonlinear system.

The computational grid is created to have fine resolution near the particle surface, where the solutions vary sharply. The logarithmic discretization is motivated
by the solution of the linearized system of Section \ref{app:linear}, 
where the largest gradients in the solution are near the particle surface.
The discretization of boundary conditions far away from the particle may be improved
by including the linear correction from the asymptotic solution, or reformulation of
far-away boundary condition according to the asymptotic behavior of the analytic solution
far away from the particle.

For strong electric field ($\beta \gg 1)$, the nonlinear effects become dominant and
a boundary-layer structure begins to form near the particle surface. 
In order to achieve high numerical accuracy, the grid discretization should be made finer
near the particle surface (at $r = 1$), while $R_{max}$ can be made smaller. Moreover, a different discretization scheme may be used for
the $r$ coordinate, to have higher grid resolution near the particles.
In addition, the continuation of the solution between different $\beta$ values
should advance in finer steps, to ensure Newton method convergence.

\section{Future Work}
Future work may include optimizing the direct sparse linear solver 
to use the ordering data from previous calls,
or replacing the full Newton step by an iterative linear solver
(e.g. using preconditioned Krylov subspace methods \cite{saad2003book} or
Multigrid), allowing finer grids to be used.
In addition, the framework may be used for automatic construction of Newton solvers 
for other nonlinear PDE systems, for example -- for particle with different surface 
properties, or different particle shape.

>>>>

The numerical solver converges well for $\beta < 5$ and the resulting steady-state velocity 
for moderate $\beta$ values is very close to being an affine function of $\beta$.
When $\beta$ becomes large, the salt concentration $C$ near the front of the particle 
(at $\theta \approx 0$) approaches zero, while creating a wake behind the particle 
(at $\theta \approx \pi$) as shown in Figures \ref{fig:LargeBeta_C1}, \ref{fig:LargeBeta_C2} and
\ref{fig:LargeBeta_C3}.
This results in high gradients in $\log C$ near the front of the particle  and, therefore, a strong electric field $\bE$ forms near the particle surface (Figures \ref{fig:LargeBeta_Phi1}, \ref{fig:LargeBeta_Phi2} and \ref{fig:LargeBeta_Phi3}) due to effective boundary conditions in (\ref{eq:ephor_bnd}). 
From Figs. \ref{fig:LargeBetaV_grids}-\ref{fig:LargeBetaRatio} we conclude that even our most accurate numerical solution is valid only up to $\beta = 5$ approximately, and therefore we only further examine solutions in this range.

The streamline pattern does not change drastically for different electric field magnitudes. 
The resulting fluid flow is nearly proportional to that of the Stokes creeping flow, 
as shown in Figure \ref{fig:LargeBeta_Psi}. 
The difference between $\Psi(r, \theta)$ and $\Psi(r, \pi-\theta)$ is shown in Figure \ref{fig:LargeBeta_PsiAsymm}, demonstrating that the flow is almost symmetric (with relative
difference up to approximately $0.5\%$) for $\beta < 5$.

The numerical results show that a boundary layer starts to form near the front 
of the particle for strong electric fields (see Figures \ref{fig:LargeBeta_BoundaryLayer_E}
and \ref{fig:LargeBeta_BoundaryLayer_C}).
The width of the boundary layer with respect to the electric field scales, to a good approximation, is: $\bar\rho(\beta) \approx 0.2 e^{-0.5 \beta}$.
This causes numerical problems for the current solver at large $\beta$, due to insufficient
grid resolution at the boundary layer of $C$ and $\bE$.

\begin{figure}
    \begin{center}
    \includegraphics[width=.45\textwidth]{figs/LargeBetaV_[128_256_512]_Rmax=10.eps}
    \includegraphics[width=.45\textwidth]{figs/LargeBetaV_[512x512]_Rmax=10_30_100.eps}
        \caption[Large $\beta$ results for $\cU$]
{Large $\beta$ results for $\cU(\beta)$ ($\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$). 
The left figure shows the numerical results of $\cU(\beta)$ for $R_{max} = 10$ and the following grid sizes: $128 \times 128$, $256 \times 256$, $512 \times 512$. 
The right figure shows the numerical results of $\cU(\beta)$ for
$512 \times 512$ grid for $R_{max} = 100, 30, 10$. For small values of $\beta$, the linear
regime of $\cU(\beta)$ is dominant. 
For moderate $\beta$ values, the steady state velocity is very close to affine function of $\beta$.
However, for larger values of $\beta$, the numerical solution
deviates significantly from the linear regime, due to convergence problems at $\beta > 6$, 
and for $\beta \sim 7$, the solver fails to converge. The numerical convergence 
depends on the grid resolution near the particle surface -- 
as it becomes finer (when $N_r$ increases or $R_{max}$ decreases), the solver convergence improves.}
	    \label{fig:LargeBetaV_grids}	    
    \end{center}
\end{figure}
\begin{figure}
    \begin{center}
    \includegraphics[width=.45\textwidth]{figs/LargeBetaV_DeltaFromLinear[128_256_512]_Rmax=10.eps}
    \includegraphics[width=.45\textwidth]{figs/LargeBetaV_DeltaFromLinear[512x512]_Rmax=10_30_100.eps}
        \caption[Large $\beta$ results for $\cU$ -- difference from linear regime]
{Large $\beta$ results for $\cU(\beta) - \beta\cU_1$ ($\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$ and $\cU_1 \approx 4.258$ is linear regime coefficient). 
The left figure shows the numerical results of $\cU(\beta)$ for $R_{max} = 10$ and the following grid sizes: $128 \times 128$, $256 \times 256$, $512 \times 512$. 
The right figure shows the numerical results of $\cU(\beta)$ for
$512 \times 512$ grid for $R_{max} = 100, 30, 10$.}
	    \label{fig:LargeBetaV_DeltaFromLinear_grids}	    
    \end{center}
\end{figure}
\begin{figure}
    \begin{center}
    \includegraphics[width=.5\textwidth]{figs/LargeBetaRatio.eps}
        \caption[Rate of convergence estimation for $\cU$ at large $\beta$]
        {Rate of convergence estimation for $\cU$ at large $\beta$  
        ($\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$), 
        based on the numerical results of $\cU(\beta)$, for the grid sizes of 
        $128 \times 128$, $256 \times 256$, $512 \times 512$. 
        The rate of convergence $r(\beta)$ is estimated by 
        $r(\beta) \triangleq \left(\cU^{4h}(\beta) - \cU^{2h}(\beta)\right)/\left(\cU^{2h}(\beta) - \cU^{h}(\beta)\right)$.
        For small $\beta$, $r(\beta) \rightarrow 4$, showing 
        quadratic convergence for the linear regime. 
        For moderate $\beta$, the convergence rate slows down, 
        and it deteriorates for large $\beta$ values.
        }
	    \label{fig:LargeBetaRatio}
    \end{center}
\end{figure}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=1\textwidth]{figs/streamlines_asymmetry.eps}
        \caption[Streamfunction asymmetry]
        {Streamline results: grid size $512 \times 512$, 
        $R_{max} = 10$, $\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$.
        The relative streamline asymmetry is computed by 
        $\|\Psi(r, \theta) - \Psi(r, \pi-\theta)\|/\|\Psi(r, \theta)\|$
        as a function of $\beta$. The results show that the streamline pattern is almost 
        symmetric, even for moderately large $\beta$ values.}
	    \label{fig:LargeBeta_PsiAsymm}	    
    \end{center}
\end{figure}

\begin{figure}
    \begin{center}
    \includegraphics[width=.45\textwidth]{figs/BoundaryLayerWidth_E1.eps}
    \includegraphics[width=.45\textwidth]{figs/BoundaryLayerWidth_E2.eps}
    \includegraphics[width=.45\textwidth]{figs/BoundaryLayerWidth_E3.eps}
    \includegraphics[width=.45\textwidth]{figs/BoundaryLayerWidth_E4.eps}
        \caption[Electric field boundary layer]
        {Electric field boundary layer formation: grid size $512 \times 512$, 
        $R_{max} = 30$, $\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$. The upper left
        graph shows normalized electric field radial component $E_r$ near the particle front
        ($\theta = 0$). The numerical results for $e(\rho) = E_r(r=1+\rho, \theta=0)$
        are normalized by $\hat{e}(\rho) = \left(e(\rho) - \beta\right)/\left(e(0) - \beta\right)$,
        where $e(\rho = \infty) = \beta$.
        The normalized electric field satisfies $\hat{e}(\rho) \in [0, 1]$ and is shown as function
        of $\rho = r - 1$ at the upper left graph and as function of $\rho e^{0.5 \beta}$ at
        the upper right graph.
        In order to estimate the width of the boundary layer, for each $\beta$, we compute the value 
        of $\bar\rho(\beta) = \min_\rho \{ \rho : \hat{e}(\rho) < 0.5\}$. Each such $\bar\rho(\beta)$
        is plotted at the left graph as a solid dot on the corresponding line.
        In addition, $\bar\rho(\beta)$ is plotted as a function of $\beta$ using logarithmic 
        $y$ axis at the lower left graph to show that $\bar\rho(\beta) \approx 0.2 e^{-0.5 \beta}$.
        Moreover, $E_r(\beta; r=1, \theta=0) / \beta$ grows exponentially as a function of $\beta$,
        as shown at the lower right graph.
        Thus, a boundary layer is indeed forming near the front of the particle.
        }
	    \label{fig:LargeBeta_BoundaryLayer_E}	    
    \end{center}
\end{figure}

\begin{figure}
    \begin{center}
    \includegraphics[width=.45\textwidth]{figs/boundary_layer_C1.eps}
    \includegraphics[width=.45\textwidth]{figs/boundary_layer_C2.eps}
        \caption[Ionic concentration boundary layer]
        {Ionic concentration boundary layer formation: grid size $512 \times 512$, 
        $R_{max} = 30$, $\Du = 1$, $\bar\zeta = 10$, $\alpha = 0.5$. The left
        graph shows the ionic concentration $C$ near the particle front
        ($\theta = 0$). 
        In addition, $C_\beta(r=1, \theta=0)$ is plotted as a function of $\beta$ using logarithmic 
        $y$ axis at the right graph to show that 
        $C_\beta(r=1, \theta=0) \approx 0.94 e^{-0.45 \beta}$.
        The ionic concentration becomes very close to zero for $\beta \approx 7$ at $r = 1$, 
        creating a singularity near the particle front.
        }
	    \label{fig:LargeBeta_BoundaryLayer_C}	    
    \end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\section{Discussion} \label{sec:discussion}
A numerical framework for a
nonlinear electrokinetic transport system is described.
The framework is used to construct numerical solvers 
for two electrokinetic problems: the migration of ion-exchangers
and surface-charged inert particle electrophoresis.
The numerical results for weak electric field 
exhibit a good correspondence to the asymptotic models as shown above.
Two interesting phenomena (cubic regime and downstream vortex) were predicted for the ion-exchanger
using the numerical results, and were verified using asymptotic expansion of the nonlinear system.
The electrophoresis problem was solved on a high-resolution grid to validate the asymptotic 
cubic correction of the steady-state velocity, as well as providing insights of the solution behavior 
for moderate and strong electric fields.
For strong electric field, the nonlinear effects become dominant and
a boundary-layer structure begins to form.
Due to diminishing ion concentration $C$ at the front of the particle,
a boundary layer of thickness approximately $0.2 e^{-0.5\beta}$ 
is formed with respect to the electric field $\bE$ at the front of the particle.
The value of $E_r$ at ($r = 1$, $\theta = 0$) is given approximately by 
$e^{\beta/2}$, while the value of $C(r = 1, \theta = 0)$ is given approximately 
by $e^{-\beta/2}$, and their product remains close to 1 for all $\beta$ that we could compute.
However, the velocity of the particle remains nearly linear with $\beta$
and the velocity field is very close to being symmetric.

\section{Future Work}
Future work may include using the numerical results described in this work to develop
an analytical solution for the strong field regime of the electrophoresis problem, which is 
of interest. The solver may be used to investigate how the solution
(and the evolving boundary layer, in particular) depends on the problem parameters.
Optimizing the direct sparse linear solver can be done by
using the ordering data from previous calls,
or replacing the full Newton step by an iterative linear solver
(e.g., using preconditioned Krylov subspace methods \cite{saad2003book} or
Multigrid), allowing finer grids to be used.
In order to achieve high numerical accuracy, the grid discretization should be made finer
near the particle surface (at $r = 1$), while $R_{max}$ can be made smaller. Moreover, a different discretization scheme may be used for
the $r$ coordinate, to have higher grid resolution near the particles.
In addition, the continuation of the solution between different $\beta$ values
should advance in finer steps, to ensure Newton method convergence.
The discretization of boundary conditions far away from the particle may be improved
by including the linear correction from the asymptotic solution, or reformulation of
far-away boundary condition according to the asymptotic behavior of the analytic solution
far away from the particle.
Moreover, the framework may be used for automatic construction of Newton solvers 
for other nonlinear PDE systems, for example, for particles with different surface 
properties, or different particle shape.
